\documentclass{article}
\input{../../../../../../../latex-templates/gtu-solutions-short/preamble.tex}
\input{../../../../../../../latex-templates/gtu-solutions-short/english-boxes.tex}
\input{../../../../../../../latex-templates/gtu-solutions-short/commands.tex}

\title{Communication Skills in English (4300002) - Summer 2025 Solution}
\date{May 30, 2025}

\begin{document}
\maketitle

\questionmarks{Question 1}{14}{marks}

\textbf{Answer the following questions. (Any seven)}

\begin{enumerate}
    \item What kind of man is Jimmy?
    \item What does the poet mean by `ever -widening thought and action'?
    \item Describe ` heaven ' imagined by the poet.
    \item What promise did Jimmy and Bob make twenty years ago?
    \item Describe the end of the story `After Twenty Years'.
    \item What did the narrator see and feel when reached atop Pari Tibba?
    \item What made the narrator feel that the leopard was following him around? Was it true that the leopard was following him?
    \item Where does the owner of the woods live?
    \item Why do you think the poet stopped in the woods?
    \item What kind of freedom does the poet desire for the country?
\end{enumerate}

\questionmarks{Question 2(a)}{5}{marks}

\textbf{Match the column `A' with column `B' (Any five)}

\textbf{column `A'}

\textbf{column `B'}

\begin{enumerate}
    \item The leopard
    \item Red colour
    \item club
    \item Gestures
    \item Queer
    \item Strange
    \item Downy
    \item Space
    \item Proxemics
    \item Stick
    \item Kinesics
    \item Soft
    \item crimson
    \item Ruskin Bond
\end{enumerate}

\questionmarks{Question 2(b)}{6}{marks}

\textbf{Write short note. (Any Two)}

\begin{enumerate}
    \item Write a note on Communication Barriers.
    \item Discuss the Communication cycle.
    \item Explain different Flows of communication.
\end{enumerate}

\questionmarks{Question 2(c)}{3}{marks}

\textbf{Choose the correct option.(Any Three)}

\begin{enumerate}
    \item Communication without words is \underline{\hspace{2cm}} communication.
    
    ( A . non verbal B . verbal C . written)
    
    \item E mail, SMS , letters are examples of \underline{\hspace{2cm}}communication.
    
    ( A . oral B . written C . non-verbal)
    
    \item The response to a sender's message is called\underline{\hspace{2cm}}.
    
    ( A . receiver B . channel C . feedback)
    
    \item The factors that make communication slow or ineffective are called\underline{\hspace{2cm}}.
    
    ( A . paralanguage B . barriers C . feedback)
\end{enumerate}

\questionmarks{Question 3(a)}{4}{marks}

\textbf{Fill in the blanks using proper modal auxiliary. (Any four)}

\begin{enumerate}
    \item I \underline{\hspace{2cm}} read English. (can, may, should)
    \item \underline{\hspace{2cm}} you live long! (Can, Would, May)
    \item We \underline{\hspace{2cm}} help the needy. (must, should, may)
    \item We \underline{\hspace{2cm}} be careful while driving. (may, can, must)
    \item My mother \underline{\hspace{2cm}} run fast when she was in school. (can, could, might)
\end{enumerate}

\questionmarks{Question 3(b)}{4}{marks}

\textbf{Fill in the blanks using appropriate form of the verb. (Any four)}

\begin{enumerate}
    \item \underline{\hspace{2cm}} you ever \underline{\hspace{2cm}} English drama? (see)
    \item The lights went off, while they \underline{\hspace{2cm}} their dinner. (have)
    \item Listen , someone \underline{\hspace{2cm}}you outside.(call)
    \item We \underline{\hspace{2cm}} for two hours. (wait)
    \item He \underline{\hspace{2cm}} the zoo next Sunday. (visit)
\end{enumerate}

\questionmarks{Question 3(c)}{6}{marks}

\textbf{Do as directed. (Any Six)}

\begin{enumerate}
    \item My brother writes clearly. (Identify the Adverb)
    \item Wow! What a scene. (Identify the interjections)
    \item Shyam is my brother. \underline{\hspace{2cm}} study in the same class. (Use the Appropriate Pronoun)
    \item He is \underline{\hspace{2cm}} honest man. (Apply a suitable determiner from: a, an, the)
    \item There \underline{\hspace{2cm}} a bunch of grapes in the basket. (Apply a suitable verb from : are, is, has)
    \item Neither of the girls play well. (spot the error and rewrite the sentence)
    \item There are book in the bag. (spot the error and rewrite the sentence)
\end{enumerate}

\questionmarks{Question 4(a)}{4}{marks}

\textbf{Join these sentences using conjunctions given in the bracket.(Any Four)}

\begin{enumerate}
    \item The boys are afraid to go near that place. They have heard that it is haunted. (because)
    \item We were walking to the other side of the island. We found a small stream.(when)
    \item The girl did not buy that expensive pair of shoes. She liked them very much. (though)
    \item I wanted to go with them very badly. My father would not let me go.(but)
    \item He walked out of the room very angrily. He slammed the door hard.(and)
\end{enumerate}

\questionmarks{Question 4(b)}{4}{marks}

\textbf{Match the sentences to the patterns.}

\begin{tabulary}{\linewidth}{L L}
\toprule
\textbf{A} & \textbf{B} \\
\midrule
SVO & The wall collapsed. \\
SVC & Ram is a boy. \\
SV & She works sincerely. \\
SVA & Heema reads a news paper. \\
\bottomrule
\end{tabulary}

\questionmarks{Question 4(c)}{3}{marks}

\textbf{Go through the sentences and change their tenses as directed. (Any Three)}

\begin{enumerate}
    \item The boy speaks the truth. (Change into Present Continuous Tense )
    \item I know this. (Change into Simple Past )
    \item Meera ate all the apples. (Change into Present Perfect Tense )
    \item I have finished my assignment. (Change into Past Perfect Tense )
\end{enumerate}

\questionmarks{Question 4(d)}{3}{marks}

\textbf{Use appropriate preposition from the bracket. (Any Three)}

\begin{enumerate}
    \item The cup is \underline{\hspace{2cm}} the cupboard.
    \item Take your legs \underline{\hspace{2cm}} the table.
    \item There are clouds \underline{\hspace{2cm}} the town.
    \item Tom sits \underline{\hspace{2cm}} Lucy and James.
\end{enumerate}

(in, over, between, off, under, among, near)

\questionmarks{Question 5(a)}{7}{marks}

\textbf{On behalf of the Principal, XYZ Polytechnic College, Parimal Cross Road, Ahmadabad, draft a letter to place an order for purchasing stationery items for your college store to the Manager, A to Z Stationery Suppliers, M. G. Road, Ahmadabad.}

\questionmarks{Question 5(b)}{7}{marks}

\textbf{Write an email to your HR head, recommending a friend for a vacant post.}

\orquestionmarks{Question 5(a)}{7}{marks}

\textbf{Pooja Patel, Electrical department, ABC Engineering College, Gandhinagar wants to join internship program in Bharat Electric Company. Write a request letter to the CEO, Bharat Electric Company, Hyderabad on her behalf.}

\orquestionmarks{Question 5(b)}{7}{marks}

\textbf{Write an e-mail to your boss using the below stated lines:}

Learnt that boss is taking up a new project- planned for investment of millions of dollars in IT unfortunately plan will not work- competitor has similar project- this project was a failure- no demand in the market- hence should stop this idea immediately.

\end{document}
